---
title: "Xgboost"
output: html_document
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Packages
```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
#install.packages("doParallel") #new pacakage
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(caret)       # Other great library for Machine Learning 
```

#Load data

```{r}
training <- read.csv("../data/training/training_fe.csv") %>%
  janitor::clean_names()

```

#Pre-processing

```{r}
set.seed(931)
training_split <- initial_split(training, 
  prop = .7, 
  strata = yield_mg_ha  # Stratify by target variable
  )

training_split
```

##data split

```{r}
data_train <- training(training_split)  # 70% of data
data_train #Training data frame
```


```{r}
data_test <- testing(training_split)  # 30% of data
data_test #Testing data frame
```


##Distribution of target variable
```{r}
ggplot() +
  geom_density(data = data_train, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = data_test, 
               aes(x = yield_mg_ha),
               color = "blue")
```

#removing raw dates before recipe
```{r}
data_train <- data_train %>% 
  select(-date_planted, -date_harvested)

data_test <- data_test %>% 
  select(-date_planted, -date_harvested)
```


##Creating Recipe
```{r}
xgb_recipe <- recipe(yield_mg_ha~ ., data = data_train) %>%
  step_novel(all_nominal_predictors()) %>% 
  step_dummy(all_nominal_predictors()) %>%    # convert factors to numeric
  step_zv(all_predictors())%>%
  step_normalize(all_numeric_predictors())


xgb_recipe

```


#Training model
##Model Specification

```{r}
xgb_model <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")  

xgb_model
```
This is the initial cross validation. The alternatives codes are below incase the model tuning run slow for you.
##Cross Validation Setup
```{r}
set.seed(250)
# Create 10-fold cross-validation resampling object from training data
resampling_foldcv <- vfold_cv(data_train, 
                              v = 10)

resampling_foldcv
resampling_foldcv$splits[[1]]
```

##Hyperparameter grid with Latin Hypercube Sampling
```{r}
xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), data_train),
  trees(),
  size = 100
)

xgb_grid
```


```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate),
                 size = trees),
             alpha = .5,
             show.legend = FALSE)
```

#Model Tuning
```{r}
set.seed(7644)
registerDoParallel(cores = parallel::detectCores() -1)

xgb_result <- tune_race_anova(object = xgb_model,
                      preprocessor = xgb_recipe,
                      resamples = resampling_foldcv,
                      grid = xgb_grid,
                      control = control_race(save_pred = TRUE))

stopImplicitCluster()


beepr::beep()
xgb_result$.metrics[[2]]
```

Specified the alternatives codes below to tune the model faster
#Alternative cross validation
```{r}
set.seed(250)
# Create 5-fold cross-validation resampling object from training data
resampling_foldcv2 <- vfold_cv(data_train, 
                              v = 5)

resampling_foldcv2
resampling_foldcv2$splits[[1]]
```


#Alternative hyperparameter
```{r}
xgb_grid2 <- grid_space_filling(
  tree_depth(range = c(3, 8)),
  min_n(range = c(5, 40)),
  learn_rate(range = c(0.01, 0.2)),
  loss_reduction(),
  sample_size = sample_prop(range = c(0.5, 0.9)),
  finalize(mtry(), data_train),
  trees(range = c(200, 800)),
  size = 30
)

```

#Faster parallel setting and racing control
```{r}
registerDoParallel(cores = 4)

ctrl <- control_race(
  save_pred = FALSE,
  verbose_elim = TRUE,
  parallel_over = "resamples"
)


```

#Alternative model tuning
```{r}
set.seed(7645)

xgb_result2 <- tune_race_anova(
  object = xgb_model,
  preprocessor = xgb_recipe,
  resamples = resampling_foldcv2,
  grid = xgb_grid2,
  control = ctrl
)

stopImplicitCluster()
beepr::beep()

```


#Review tuning results

```{r}
collect_metrics(xgb_result)
```

#Select best models
```{r}
# Based on lowest RMSE
best_rmse <- xgb_result %>% 
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```

```{r}
# Based on lowers RMSE within 1% loss
best_rmse_pct_loss <- xgb_result %>% 
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1
                     )%>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss

```


```{r}
# Based on lowest RMSE within 1 se
best_rmse_one_std_err <- xgb_result %>% 
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

```{r}
# Based on greatest R2
best_r2 <- xgb_result %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```


```{r}
# Based on lowers R2 within 1% loss
best_r2_pct_loss <- xgb_result %>% 
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```

```{r}
# Based on lowest R2 within 1 se
best_r2_one_std_error <- xgb_result %>% 
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

#Comparing and finalizing model
```{r}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)

```


#Final Specification
```{r}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

```{r}
knitr::purl("04_xgboost.rmd", output = "04_xgboost_script.R", documentation = 0)
```

