---
title: "05-2.2_Cubist"
output: html_document
date: "2026-04-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 05_cubist_model

# Tuning: full sample, 10-fold CV, regular grid search
# Evaluation: 30% holdout
# Final model: refit on full 2014-2023 data for 2024 predictions

```{r}

library(tidyverse)
library(tidymodels)
library(rules)
library(Cubist)
library(doParallel)
```

# Load data 
```{r}

training <- read_csv("../data/training/training_fe.csv")
```

# Prepare modeling dataset 
```{r}

training_model <- training %>%
  mutate(
    previous_crop = as.factor(previous_crop),
    year          = as.integer(year)
  ) %>%
  drop_na()

cat("Total rows for modeling:", nrow(training_model), "\n")
```

# Stratified 70/30 split by yield 
```{r}

set.seed(931)
yield_split <- initial_split(training_model, prop = 0.70, strata = yield_mg_ha)

train_data <- training(yield_split)
test_data  <- testing(yield_split)

cat("Train rows:", nrow(train_data), "\n")
cat("Test rows:",  nrow(test_data),  "\n")

train_tune <- train_data
```

# Recipe 
```{r}

# Cubist handles categoricals natively — no dummy encoding needed
cubist_recipe <- recipe(yield_mg_ha ~ ., data = train_tune) %>%
  step_rm(date_planted, date_harvested) %>%
  step_zv(all_predictors())

# verify levels after prep
cat("Hybrid levels after step_other:",
    cubist_recipe %>%
      prep(training = train_tune) %>%
      bake(new_data = NULL) %>%
      pull(hybrid) %>%
      n_distinct(), "\n")
```

# Model spec
```{r}

cubist_spec <- cubist_rules(
  committees = tune(),
  neighbors  = tune()
) %>%
  set_engine("Cubist") %>%
  set_mode("regression")
```

# Workflow 
```{r}

cubist_wf <- workflow() %>%
  add_recipe(cubist_recipe) %>%
  add_model(cubist_spec)
```

# 10-fold CV on subsample stratified by yield 
```{r}

set.seed(123)
folds <- vfold_cv(train_tune, v = 10, strata = yield_mg_ha)
```

# Regular grid 
```{r}

cubist_grid <- grid_regular(
  committees(range = c(1, 20)),
  neighbors(range  = c(0, 9)),
  levels = 5
)

cat("Grid size:", nrow(cubist_grid), "combinations\n")
```

# Parallel backend 
```{r}
n_cores <- parallel::detectCores() - 1
cat("Using", n_cores, "cores\n")
registerDoParallel(cores = n_cores)
```

# Tune 
```{r}

cat("Starting tuning...\n")

cubist_tune <- tune_grid(
  cubist_wf,
  resamples = folds,
  grid      = cubist_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(verbose = TRUE, save_pred = TRUE)
)

stopImplicitCluster()
```

# Tuning results 
```{r}

cat("Tuning complete. Best results:\n")
show_best(cubist_tune, metric = "rmse")

autoplot(cubist_tune)
ggsave("../output/cubist_tuning_plot.png", width = 8, height = 5, dpi = 300)
```

# Select best parameters 
```{r}

best_params <- select_best(cubist_tune, metric = "rmse")
cat("Best parameters:\n")
print(best_params)
```

## Metrics on train dataset
```{r}
train_preds <- predict(eval_fit, train_data) %>%
  bind_cols(train_data %>% select(yield_mg_ha))

train_metrics <- train_preds %>% 
  metrics(truth = yield_mg_ha, estimate = .pred)

print(train_metrics)
```


# Evaluate on 30% holdout 
```{r}

# fit on full train_data with best params for honest evaluation
eval_recipe <- recipe(yield_mg_ha ~ ., data = train_data) %>%
  step_rm(date_planted, date_harvested) %>%
  step_other(hybrid, threshold = 0.001) %>%
  step_zv(all_predictors())

eval_wf <- workflow() %>%
  add_recipe(eval_recipe) %>%
  add_model(finalize_model(cubist_spec, best_params))

cat("Fitting evaluation model on train_data...\n")
eval_fit <- fit(eval_wf, data = train_data)

test_preds <- predict(eval_fit, test_data) %>%
  bind_cols(test_data %>% select(yield_mg_ha))

test_metrics <- test_preds %>%
  metrics(truth = yield_mg_ha, estimate = .pred)

cat("Holdout Test Performance:\n")
print(test_metrics)
```


# Variable Importance
```{r}
# Variable Importance
library(vip)

cubist_vip <- eval_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 20)

cubist_vip

ggsave("../output/cubist_vip.png", width = 8, height = 5, dpi = 300)

# save importance scores as dataframe too
cubist_importance <- eval_fit %>%
  extract_fit_parsnip() %>%
  vi()

write_csv(cubist_importance, "../output/cubist_importance.csv")
```


## Residual plots
```{r}

#Residual by site
test_preds %>% 
  bind_cols(test_data %>% select(site, year)) %>% 
  mutate(residual = yield_mg_ha - .pred) %>% 
  group_by(site) %>%
  summarise(mean_residual = mean(residual), rmse = sqrt(mean(residual ^2))) %>%
  write.csv("../output/cubist_residuals_by_site.csv")

#Residual by year
test_preds %>% 
  bind_cols(test_data %>% select(site, year)) %>% 
  mutate(residual = yield_mg_ha - .pred) %>% 
  group_by(year) %>%
  summarise(mean_residual = mean(residual), rmse = sqrt(mean(residual ^2))) %>%
  write.csv("../output/cubist_residuals_by_year.csv")

#Full test predictions with site and year for Shiny
test_preds %>%
  bind_cols(test_data %>% select(site, year, hybrid)) %>% 
  write.csv("../output/cubist_test_predictions.csv")

#Full train predictions with site and year for Shiny
predict(eval_fit, train_data) %>%
  bind_cols(train_data %>% select(yield_mg_ha, site, year, hybrid)) %>% 
  write.csv("../output/cubist_train_predictions.csv")

```


# Predicted vs actual plot 
```{r}

test_preds %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(color = "steelblue",size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "darkred", linetype = "dashed") +
  labs(title = "Cubist: Predicted vs Actual Yield",
       x = "Actual Yield (Mg/ha)",
       y = "Predicted Yield (Mg/ha)") +
  theme_minimal()

ggsave("../output/cubist_predicted_vs_actual.png", width = 8, height = 5, dpi = 300)

```

# Refit on FULL 2014-2023 data for 2024 predictions 
```{r}

cat("Refitting final model on full 2014-2023 data...\n")

final_recipe <- recipe(yield_mg_ha ~ ., data = training_model) %>%
  step_rm(date_planted, date_harvested) %>%
  step_other(hybrid, threshold = 0.001) %>%
  step_zv(all_predictors())

final_cubist_wf <- workflow() %>%
  add_recipe(final_recipe) %>%
  add_model(finalize_model(cubist_spec, best_params))

final_cubist_fit <- fit(final_cubist_wf, data = training_model)

```

## 2024 Predictions
```{r}

testing_pred <- testing %>%
  mutate(
    replicate = NA_real_, 
    block = NA_real_, 
    grain_moisture = NA_real_, 
    date_planted = as.Date(NA), 
    date_harvested = as.Date(NA), 
    year_factor = NA_real_, 
    doy_planted = NA_real_, 
    doy_harvested = NA_real_,
    season_length = NA_real_
  )

#predict using final model 
predictions_2024 <- predict(final_cubist_fit, testing_pred) %>%
  bind_cols(testing %>% select(year, site, hybrid))

head(predictions_2024)
sum(is.na(predictions_2024$.pred))

write_csv(predictions_2024, "../output/testing_final_pred.csv")

```



# Save everything
```{r}

saveRDS(final_cubist_fit, "../output/cubist_final_model.rds")
saveRDS(eval_fit,         "../output/cubist_eval_model.rds")
saveRDS(cubist_tune,      "../output/cubist_tune_results.rds")
write_csv(test_metrics,   "../output/cubist_test_metrics.csv")
write_csv(train_metrics,   "../output/cubist_train_metrics.csv")

cat("Cubist script complete\n")

```


From, variable importance grain mositure is imp variable which is not available in testing dataset. So, trying to predict without grain mositure. 


## Running without grain moisture

```{r}
# Define best params directly — no need to reload cubist_tune
best_params <- tibble(committees = 20, neighbors = 9)

# Refit final model WITHOUT grain_moisture
final_recipe_no_gm <- recipe(yield_mg_ha ~ ., data = training_model) %>%
  step_rm(date_planted, date_harvested, grain_moisture) %>%
  step_zv(all_predictors())

final_cubist_wf_no_gm <- workflow() %>%
  add_recipe(final_recipe_no_gm) %>%
  add_model(finalize_model(cubist_spec, best_params))

final_cubist_fit_no_gm <- fit(final_cubist_wf_no_gm, data = training_model)

```


```{r}
# Reload testing data
testing <- read_csv("../data/testing/testing_final.csv") %>%
  mutate(
    previous_crop = as.factor(previous_crop),
    year          = as.integer(year)
  )

# Add missing columns that training had but testing doesn't
testing_pred <- testing %>%
  mutate(
    replicate      = NA_real_,
    block          = NA_real_,
    grain_moisture = NA_real_,
    date_planted   = as.Date(NA),
    date_harvested = as.Date(NA),
    year_factor    = as.numeric(as.character(year)),
    doy_planted    = NA_real_,
    doy_harvested  = NA_real_,
    season_length  = NA_real_,
    previous_crop  = as.character(previous_crop),
    previous_crop  = tolower(str_trim(previous_crop)),
    previous_crop  = case_when(
      str_detect(previous_crop, "soybean|soya") ~ "soybean",
      str_detect(previous_crop, "corn")         ~ "corn",
      str_detect(previous_crop, "wheat")        ~ "wheat",
      str_detect(previous_crop, "sorghum")      ~ "sorghum",
      str_detect(previous_crop, "cotton")       ~ "cotton",
      str_detect(previous_crop, "peanut")       ~ "peanut",
      str_detect(previous_crop, "clover")       ~ "other",
      is.na(previous_crop)                      ~ "unknown",
      TRUE                                      ~ previous_crop
    ),
    previous_crop = as.factor(previous_crop)
  )

predictions_2024 <- predict(final_cubist_fit_no_gm, testing_pred) %>%
  bind_cols(testing %>% select(year, site, hybrid))

cat("Mean:", mean(predictions_2024$.pred), "\n")
cat("SD:",   sd(predictions_2024$.pred), "\n")
cat("NAs:",  sum(is.na(predictions_2024$.pred)), "\n")

write_csv(predictions_2024, "../output/testing_final_pred_Cubist_no_gm.csv")
```

## Submission file
```{r}

submission <- read_csv("../data/testing/testing_submission.csv")

submission_filled <- submission %>%
  left_join(predictions_2024 %>% select(year, site, hybrid, .pred),
            by = c("year", "site", "hybrid")) %>%
  mutate(yield_mg_ha = .pred) %>%
  select(-`.pred`)

sum(is.na(submission_filled$yield_mg_ha))
write_csv(submission_filled, "../output/testing_submission_final.csv")

```

