---
title: "05-2.2_Cubist"
output: html_document
date: "2026-04-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 05_cubist_model

# Tuning: 20k subsample, 10-fold CV, regular grid search
# Evaluation: 30% holdout
# Final model: refit on full 2014-2023 data for 2024 predictions

```{r}

library(tidyverse)
library(tidymodels)
library(rules)
library(Cubist)
library(doParallel)
```

# Load data 
```{r}

training <- read_csv("../data/training/training_fe.csv")
```

# Prepare modeling dataset 
```{r}

training_model <- training %>%
  mutate(
    previous_crop = as.factor(previous_crop),
    year          = as.integer(year)
  ) %>%
  drop_na()

cat("Total rows for modeling:", nrow(training_model), "\n")
```

# Stratified 70/30 split by yield 
```{r}

set.seed(931)
yield_split <- initial_split(training_model, prop = 0.70, strata = yield_mg_ha)

train_data <- training(yield_split)
test_data  <- testing(yield_split)

cat("Train rows:", nrow(train_data), "\n")
cat("Test rows:",  nrow(test_data),  "\n")

train_tune <- train_data
```

# Recipe 
```{r}

# Cubist handles categoricals natively — no dummy encoding needed
cubist_recipe <- recipe(yield_mg_ha ~ ., data = train_tune) %>%
  step_rm(date_planted, date_harvested) %>%
  step_zv(all_predictors())

# verify levels after prep
cat("Hybrid levels after step_other:",
    cubist_recipe %>%
      prep(training = train_tune) %>%
      bake(new_data = NULL) %>%
      pull(hybrid) %>%
      n_distinct(), "\n")
```

# Model spec
```{r}

cubist_spec <- cubist_rules(
  committees = tune(),
  neighbors  = tune()
) %>%
  set_engine("Cubist") %>%
  set_mode("regression")
```

# Workflow 
```{r}

cubist_wf <- workflow() %>%
  add_recipe(cubist_recipe) %>%
  add_model(cubist_spec)
```

# 10-fold CV on subsample stratified by yield 
```{r}

set.seed(123)
folds <- vfold_cv(train_tune, v = 10, strata = yield_mg_ha)
```

# Regular grid 
```{r}

cubist_grid <- grid_regular(
  committees(range = c(1, 20)),
  neighbors(range  = c(0, 9)),
  levels = 5
)

cat("Grid size:", nrow(cubist_grid), "combinations\n")
```

# Parallel backend 
```{r}
n_cores <- parallel::detectCores() - 1
cat("Using", n_cores, "cores\n")
registerDoParallel(cores = n_cores)
```

# Tune 
```{r}

cat("Starting tuning...\n")

cubist_tune <- tune_grid(
  cubist_wf,
  resamples = folds,
  grid      = cubist_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(verbose = TRUE, save_pred = TRUE)
)

stopImplicitCluster()
```

# Tuning results 
```{r}

cat("Tuning complete. Best results:\n")
show_best(cubist_tune, metric = "rmse")

autoplot(cubist_tune)
ggsave("../output/cubist_tuning_plot.png", width = 8, height = 5, dpi = 300)
```

# Select best parameters 
```{r}

best_params <- select_best(cubist_tune, metric = "rmse")
cat("Best parameters:\n")
print(best_params)
```

## Metrics on train dataset
```{r}
train_preds <- predict(eval_fit, train_data) %>%
  bind_cols(train_data %>% select(yield_mg_ha))

train_metrics <- train_preds %>% 
  metrics(truth = yield_mg_ha, estimate = .pred)

print(train_metrics)
```


# Evaluate on 30% holdout 
```{r}

# fit on full train_data with best params for honest evaluation
eval_recipe <- recipe(yield_mg_ha ~ ., data = train_data) %>%
  step_rm(date_planted, date_harvested) %>%
  step_other(hybrid, threshold = 0.001) %>%
  step_zv(all_predictors())

eval_wf <- workflow() %>%
  add_recipe(eval_recipe) %>%
  add_model(finalize_model(cubist_spec, best_params))

cat("Fitting evaluation model on train_data...\n")
eval_fit <- fit(eval_wf, data = train_data)

test_preds <- predict(eval_fit, test_data) %>%
  bind_cols(test_data %>% select(yield_mg_ha))

test_metrics <- test_preds %>%
  metrics(truth = yield_mg_ha, estimate = .pred)

cat("Holdout Test Performance:\n")
print(test_metrics)
```


# Variable Importance
```{r}
# Variable Importance
library(vip)

cubist_vip <- eval_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 20)

cubist_vip

ggsave("../output/cubist_vip.png", width = 8, height = 5, dpi = 300)

# save importance scores as dataframe too
cubist_importance <- eval_fit %>%
  extract_fit_parsnip() %>%
  vi()

write_csv(cubist_importance, "../output/cubist_importance.csv")
```


# Predicted vs actual plot 
```{r}

test_preds %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(color = "steelblue",size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "darkred", linetype = "dashed") +
  labs(title = "Cubist: Predicted vs Actual Yield",
       x = "Actual Yield (Mg/ha)",
       y = "Predicted Yield (Mg/ha)") +
  theme_minimal()

ggsave("../output/cubist_predicted_vs_actual.png", width = 8, height = 5, dpi = 300)
```

# Refit on FULL 2014-2023 data for 2024 predictions 
```{r}

cat("Refitting final model on full 2014-2023 data...\n")

final_recipe <- recipe(yield_mg_ha ~ ., data = training_model) %>%
  step_rm(date_planted, date_harvested) %>%
  step_other(hybrid, threshold = 0.001) %>%
  step_zv(all_predictors())

final_cubist_wf <- workflow() %>%
  add_recipe(final_recipe) %>%
  add_model(finalize_model(cubist_spec, best_params))

final_cubist_fit <- fit(final_cubist_wf, data = training_model)
```

# Save everything
```{r}
saveRDS(final_cubist_fit, "../output/cubist_final_model.rds")
saveRDS(eval_fit,         "../output/cubist_eval_model.rds")
saveRDS(cubist_tune,      "../output/cubist_tune_results.rds")
write_csv(test_metrics,   "../output/cubist_test_metrics.csv")
write_csv(train_metrics,   "../output/cubist_train_metrics.csv")

cat("Cubist script complete\n")
```

