---
title: "xgboost"
output: html_document
date: "2026-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Packages
```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
#install.packages("doParallel") #new pacakage
library(tidymodels)   
library(finetune)     
library(vip)         
library(xgboost)      
library(ranger)       
library(tidyverse)  
library(doParallel)   
library(caret)    
library(yardstick)
```


#Load data

```{r}
training <- read.csv("../data/training/training_fe.csv") %>%
  janitor::clean_names()

training

```

#Pre-processing

```{r}
set.seed(931)
training_split <- initial_split(training, 
  prop = .7, 
  strata = yield_mg_ha
  )

training_split
```

#Data Splitting
##Training Data
```{r}
data_train <- training(training_split)
data_train 
```

##Test Data
```{r}
data_test <- testing(training_split)  # 30% of data
data_test #Testing data frame
```

#Remove Raw Date Variables
```{r}
data_train2 <- data_train %>% 
  select(-date_planted, -date_harvested)

data_test2 <- data_test %>% 
  select(-date_planted, -date_harvested)
```


#Recipe Creation
```{r}
# Recipe
xgb_recipe <- recipe(yield_mg_ha ~ ., data = data_train2) %>%
  step_novel(all_nominal_predictors()) %>%                  
  step_other(hybrid, threshold = 0.01) %>%              
  step_zv(all_predictors()) %>%                           
  step_dummy(all_nominal_predictors(), sparse = "yes")


xgb_recipe
```


#Model Specification
```{r}
xgb_model <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  stop_iter = 25
) %>%
  set_engine(
    "xgboost",
    validation = 0.1,
    counts = FALSE,
    verbose = 0
  ) %>%
  set_mode("regression")

xgb_model
```

#Workflow Setup
```{r}
# Workflow
xgb_wf <- workflow() %>%
  add_recipe(xgb_recipe) %>%
  add_model(xgb_model)

xgb_wf

```

#3-Fold Cross-Validation Setup
```{r}
set.seed(250)
resampling_foldcv <- vfold_cv(data_train2, v = 3)

resampling_foldcv
```

#Hyperparameter Grid
```{r}
# Parameter set
xgb_params <- extract_parameter_set_dials(xgb_wf) %>%
  update(
    mtry = mtry_prop(range = c(0.05, 0.50)),
trees = trees(range = c(200L, 800L))
  )

xgb_params

# Grid
set.seed(250)
xgb_grid <- grid_space_filling(
  xgb_params,
  size = 20
)

xgb_grid
```

#Model Tuning
```{r}
# Parallel setup
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

# Tune model
set.seed(7644)

xgb_result <- tune_race_anova(
  object = xgb_wf,
  resamples = resampling_foldcv,
  grid = xgb_grid,
  control = control_race(
    burn_in = 2,
    save_pred = FALSE,
    verbose_elim = TRUE
  )
)

# Stop parallel
stopCluster(cl)
registerDoSEQ()

# Alert
beepr::beep()
```


#Checking tuning results
```{r}
show_notes(xgb_result)
```

#Racing Plot
```{r}
plot_race(xgb_result)

ggsave("../output/xgboost_tuning_plot.png", width = 8, height = 5, dpi = 300)
```

#Tuning Metrics
```{r}
collect_metrics(xgb_result)
```

#Best RMSE Selection
```{r}
best_rmse <- xgb_result %>% 
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```


#Based on lowest RMSE within 1% loss
```{r}
best_rmse_pct_loss <- xgb_result %>% 
  select_by_pct_loss(
    metric = "rmse",
    limit = 1,
    min_n
  ) %>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```

#Based on lowest RMSE within 1 se
```{r}
best_rmse_one_std_err <- xgb_result %>% 
  select_by_one_std_err(
    metric = "rmse",
    desc(trees)
  ) %>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

#Best R-Squared Selection
```{r}
best_r2 <- xgb_result %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

#Based on lowers R2 within 1% loss
```{r}
best_r2_pct_loss <- xgb_result %>% 
  select_by_pct_loss(
    metric = "rsq",
    limit = 1,
    min_n
  ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```


#Based on lowest R2 within 1 se
```{r}
best_r2_one_std_error <- xgb_result %>% 
  select_by_one_std_err(
    metric = "rsq",
    desc(trees)
  ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

#Model Comparison
```{r}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)

```

#Final Model Evaluation(Test Set)
```{r}
my_metrics <- metric_set(rmse, rsq, mae)
best_rmse <- select_best(xgb_result, 
                         metric = "rmse")

final_wf <- finalize_workflow(xgb_wf, 
                              best_rmse)

final_fit <- last_fit(final_wf, 
                      split = training_split,
                      metrics = my_metrics
                      )

collect_metrics(final_fit)

collect_predictions(final_fit)

test_metrics <- collect_metrics(final_fit)

write_csv(
  test_metrics,
  "../output/xgboost_test_metrics.csv"
)
```

#Predicted vs Observed Plot
```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(alpha = 0.4) +
  geom_abline(color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue") +
  coord_equal() +
  labs(title = "Xgboost: Predicted vs Actual Yield",
       x = "Observed Yield (Mg/ha)",
       y = "Predicted Yield (Mg/ha)")

ggsave("../output/xgboost_predicted_vs_actual_yield.png", width = 8, height = 5, dpi = 300)
```


#Training Set Performance
```{r}
# fit on full training data
final_model <- fit(final_wf, data = data_train)

# training set predictions
train_preds <- predict(final_model, new_data = data_train) %>%
  bind_cols(data_train %>%
              select(yield_mg_ha))

# training RMSE, R2, and MAE
train_preds %>%
  metrics(truth = yield_mg_ha, estimate = .pred)

train_preds <- collect_metrics(final_fit)

write_csv(
  train_preds,
  "../output/xgboost_train_metrics.csv"
)
```



#Variable Importance
```{r}
xgb_fit <- final_model %>%
  extract_fit_parsnip()

vip(xgb_fit) +
  theme_minimal()

ggsave("../output/xgboost_vip.png", width = 8, height = 5, dpi = 300)

# save important variables as new dataframe
xgboost_importance <- xgb_fit %>%
  vi()

write_csv(xgboost_importance, "../output/xgboost_importance.csv")

```
##2024 Test Set Prediction
#Loading 2024 test set
```{r}
testing_2024<- read.csv("../data/testing/testing_final.csv")
testing_2024
```

#Removing unavailable variables in test set 

```{r}
missing_2024_cols <- c(
  "replicate",
  "block",
  "grain_moisture",
  "soilp_h",
  "year_factor",
  "doy_planted",
  "doy_harvested",
  "season_length",
  "date_planted",
  "date_harvested"
)

training_2024_model <- training %>%
  select(-any_of(missing_2024_cols))

testing_2024_model <- testing_2024 %>%
  select(-any_of(missing_2024_cols))

```

#Recreating recipe
```{r}
xgb_recipe_2024 <- recipe(yield_mg_ha ~ ., data = training_2024_model) %>%
  step_novel(all_nominal_predictors()) %>%
  step_other(hybrid, threshold = 0.01) %>%
  step_zv(all_predictors()) %>%
  step_dummy(all_nominal_predictors(), sparse = "yes")
```

#Finalizing Workflow
```{r}
xgb_wf_2024 <- workflow() %>%
  add_recipe(xgb_recipe_2024) %>%
  add_model(xgb_model)

best_rmse <- select_best(xgb_result, metric = "rmse")

final_wf_2024 <- finalize_workflow(
  xgb_wf_2024,
  best_rmse
)
```

#Refit model on FULL 2014-2024 data
```{r}
final_fit_2024 <- fit(
  final_wf_2024,
  data = training_2024_model
  )
```




#2024 Prediction
```{r}
predictions_2024 <- predict(
  final_fit_2024,
  new_data = testing_2024_model
) %>%
  bind_cols(testing_2024 %>% select(year, site, hybrid))

head(predictions_2024)

sum(is.na(predictions_2024$.pred))

write_csv(
  predictions_2024,
  "../output/testing_final_pred_xgboost.csv"
)
```



